Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-04 paper/paper.tex
many changes in progress

\newtoggle{full}
\togglefalse{full}

@@

On top of these two approaches are methods in which a new
language is implemented `inside' another language of the same
level of abstraction. 
}{%
One way to implement a programming language is to embed it within another language.
}%
Different languages provide different
means for this; Lisp is famous for the macro facility
that allows to extend the language almost without
restriction --- by writing \textit{macros}, one adds new
constructs to the existing language. There are several uses of
macros --- one is to extend the language \textit{syntax}, for
example, by adding new control structures; another is to keep
the existing syntax but alter the \textit{operational semantics}
--- the way programs are executed and compute their outputs.

@@

@@

Anglican is intended to co-exist with Clojure and be a part of
the source of a Clojure program. To facilitate this, Anglican
programs, or queries, are wrapped by macros (defined in the
\texttt{anglican.emit} namespace), which call the CPS
transformations and define Clojure values
%Clojure objects
suitable for passing
as arguments to inference algorithms (\texttt{defquery},
\texttt{query}). In addition to defining entire queries,
Anglican promotes modularization of inference algorithms through
the definitions of \textit{probabilistic functions} using
\texttt{defm}
%and \texttt{fm}
(Anglican counterparts of Clojure \texttt{defn}).
%and \texttt{fn}).
Probabilistic functions are
written in Anglican, may include probabilistic forms
\texttt{sample} and \texttt{observe}, and can be seamlessly
called from inside Anglican queries, just like functions locally
defined within the same query.

@@

@@

\iftoggle{full}{%
\subsection{Expression Kinds}
\label{sec:expr}

There are three different kinds of inputs to CPS transformation:
\begin{itemize}
    \item Literals, which are constant expressions. They
       are passed as an argument to the
        continuation unmodified.
    \item Value expressions such as the \texttt{fn} form
        (called \textit{opaque}
        expressions in the code). They must be transformed to
        CPS, but the transformed object is passed to the
        continuation as a whole, opaquely.
    \item General expressions (which we call
        \textit{transparent} expressions).
        The continuation is threaded through such an expression
        in an expression-specific way, and can be
        called in multiple locations of the CPS-transformed
        code, such as in all branches of an \texttt{if}
        statement.
\end{itemize}

\subsubsection{Literals}
\label{sec:literals}

Literals are the same in Anglican and Clojure. They are
left unmodified; literals are a subset of opaque expressions.
However, the Clojure syntax has a peculiarity
of using the syntax of compound literals (vectors, hash maps,
and sets) for data constructors. Hence, compound literals must
be traversed recursively, and if there is a nested non-literal
component, transformed into a call to the corresponding data
constructor. Functions \texttt{cps-of-vector},
\texttt{cps-of-hash-map}, \texttt{cps-of-set}, called from
\texttt{cps-of-expression}, transform Clojure constructor syntax
(\texttt{[...]}, \texttt{\{...\}}, \texttt{\#\{...\}}) into the
corresponding calls:
\begin{lstlisting}[style=default]
=> (cps-of-vector [0 1 2] 'cont)
(cont (vector 0 1 2) $state)
=> (cps-of-hash-map {:a 1, :b 2} 'cont)
(cont (hash-map :a 1, :b 2) $state)
=> (cps-of-set #{0 1} 'cont)
(cont (set (list 0 1)) $state)
\end{lstlisting}

\subsubsection{Opaque Expressions}
\label{sec:opaque}

Opaque, or value, expressions, have a different shape in the
original and the CPS form. However, their CPS form follows the
pattern \texttt{(continuation transformed-expression \$state)}, and thus
the transformation does not depend on the continuation parameter, and
can be accomplished without passing the parameter as a
transformation argument. Primitive (non-CPS) procedures used in
Anglican code, \texttt{(fn ...)} forms, and \text{(mem ...)}
forms are opaque and transformed by
\texttt{primitive-}\linebreak[0]\texttt{procedure-}\linebreak[0]\texttt{cps},
\texttt{fn-cps}, and \texttt{mem-cps}, correspondingly: a
slightly simplified CPS form of expression
\begin{lstlisting}[style=default]
(fn [x y]
  (+ x y))
\end{lstlisting}
would be
\begin{lstlisting}[style=default]
(fn [cont $state x y]
  (cont (+ x y) $state))
\end{lstlisting}
In the actual code an automatically generated fresh symbol is
used instead of \texttt{cont}.

\subsubsection{General Expressions}
\label{seq:general}

The most general form of CPS transformation receives an
expression and a continuation as parameters, and returns the expression
in CPS form with the continuation parameter potentially called in multiple
tail positions. General expressions can be somewhat voluntarily
divided into several groups:
\begin{itemize}
    \item binding forms --- \texttt{let} and
        \texttt{loop/recur};
    \item flow control --- \texttt{if}, \texttt{when},
        \texttt{cond}, \texttt{case}, \texttt{and}, \texttt{or} and \texttt{do};
    \item function applications and \texttt{apply};
    \item probabilistic forms --- \texttt{observe},
        \texttt{sample}, \texttt{store}, and \texttt{retrieve}.
\end{itemize}
Functions that transform general expressions accept the
expression and the continuation as parameters, and are
consistently named \texttt{cps-of-}\textit{form}, for example,
\texttt{cps-of-do}, \texttt{cps-of-store}.

}{%
}

@@

\paragraph{Managing stack size}
}{%
Continuations may lead to unbounded stack growth in
recursive code.
}%
In implementations of functional programming languages stack
growth is avoided through \textit{tail call optimization} (TCO).
However, Clojure does not support a general form of TCO, and
CPS-transformed code that creates deeply nested calls will
easily exhaust the stack. Anglican employs a workaround called
\textit{trampolining} --- instead of inserting a continuation
call directly, the transformer always wraps the call into a
\textit{thunk}, or parameterless function. The thunk is returned
and called by the trampoline (Clojure provides function
\texttt{trampoline} for this purpose) --- this way the
computation continues, but the stack is collapsed on every
continuation call. Function \texttt{continue} implements the
wrapping and is invoked on every continuation call:
\begin{lstlisting}[style=default]
=> (continue 'cont 'value 'state)
(fn [] (cont value state))
\end{lstlisting}
Correspondingly, the full, wrapped CPS form of
\begin{lstlisting}[style=default]
(fn [x y] (+ x y))
\end{lstlisting}
becomes
\begin{lstlisting}[style=default]
(fn [cont $state x y]
  (fn []
    (cont (+ x y) $state)))
\end{lstlisting}
When the CPS-transformed function is called, it returns a
\textit{thunk} (a parameterless function) which is then
re-invoked through the trampoline, with the stack collapsed.

\iftoggle{full}{%
\paragraph{Avoiding exponential code growth}

To realize potential danger of code size explosion, consider
CPS transformation of code
\begin{lstlisting}[style=default]
(if (adult? person)
  (if (male? person)
     (choose-beer)
     (choose-wine))
  (choose-juice))
\end{lstlisting}
with continuation
\begin{lstlisting}[style=default]
(fn [choice _]
  (case (kind choice)
    :beer (beer-jar choice)
    :wine (wine-glass choice)
    :juice (juice-bottle choice)))
\end{lstlisting}
During CPS transformation, if we substitute the code of this continuation 
for all of its calls, the code will be repeated three times in 
the CPS-transformed expression. In general, CPS code can
grow extremely large if the code of continuations is substituted
repeatedly.

To circumvent this inefficiency, CPS transformers for
expressions with multiple continuation points (\texttt{if} and
derivatives, \texttt{and}, \texttt{or}, and \texttt{case}) bind
the continuation to a fresh symbol if it is not yet a symbol.
Macro \texttt{defn-with-{\linebreak[0]}named-{\linebreak[0]}cont}
establishes the binding automatically:
\begin{lstlisting}[style=default]
=> (cps-of-if '(c t f) '(fn [x] (* 2 x)))
(let [cont (fn [x] (* 2 x))]
  (if c
      (fn [] (cont t $state))
      (fn [] (cont f $state))))
\end{lstlisting}
}{
}

@@

To overcome the above problems, Anglican uses a different scheme
which is almost as efficient as the scheme based on stack addresses for
reuse of previously drawn values, while producing addresses of constant size,
and allows manual computation of checkpoint addresses at runtime
when the default automatic scheme is insufficient. According to
the scheme:
\begin{itemize}
    \item A checkpoint may accept an auxiliary argument --- the
        checkpoint identifier. If specified, the identifier is the first
        argument of a checkpoint. For example \texttt{(sample
        'x1  (normal 0 1))} defines a \texttt{sample} checkpoint
        with identifier \texttt{x1}.
    \item If the identifier is omitted, a unique identifier is
        generated automatically as a fresh symbol.
    \item At runtime, the address of a checkpoint invocation has
        the form
        \texttt{[}\textit{checkpoint-identifier
        number-of-previous-occurrences}\texttt{]}, where the
        occurrences are of a checkpoint with the same checkpoint
        identifier.
    \item If a sequence of invocations of the same checkpoint is
        interrupted by a different checkpoint, the number of
        previous occurrences is rounded up to a multiple of
        an integer. For efficiency, a small power of 2 is used,
        such as $2^4 = 16$.
\end{itemize}

@@

% \begin{itemize}
%     \item Values produced by the process are sampled from the distribution
%         returned by \texttt{produce}.
%     \item When a value is observed from the distribution, a new
%         process, with the observed value absorbed, is returned
%         by \texttt{absorb}.
% \end{itemize}

% A common pattern when programming with random processes in 
% Anglican is to generate a sequences of sampled values and 
% construct an updated process at each step.

>>>>

\newtoggle{full}
\input{toggle}

@@

On top of these two approaches are methods in which a new
language is implemented `inside' another language of the same
level of abstraction. Different languages provide different
means for this; }{%
There are several ways to build a new language on top of an
existing one;
}%
Lisp is famous for the macro facility
that allows to extend the language almost without
restriction --- by writing \textit{macros}, one adds new
constructs to the existing language. There are several uses of
macros --- one is to extend the language \textit{syntax}, for
example, by adding new control structures; another is to keep
the existing syntax but alter the \textit{operational semantics}
--- the way programs are executed and compute their outputs.

@@

\subsection*{Contributions}

@@

Anglican is intended to co-exist with Clojure and be a part of
the source of a Clojure program. To facilitate this, Anglican
programs, or queries, are wrapped by macros (defined in the
\texttt{anglican.emit} namespace), which call the CPS
transformations and define Clojure values suitable for passing
as arguments to inference algorithms (\texttt{defquery},
\texttt{query}). In addition to defining entire queries,
Anglican promotes modularization of inference algorithms through
the definitions of \textit{probabilistic functions} using
\texttt{defm}
%and \texttt{fm}
(Anglican counterparts of Clojure \texttt{defn}).
%and \texttt{fn}).
Probabilistic functions are
written in Anglican, may include probabilistic forms
\texttt{sample} and \texttt{observe}, and can be seamlessly
called from inside Anglican queries, just like functions locally
defined within the same query.

@@

\section{Probabilistic Program}
\label{sec:pp-def}

@@

\iftoggle{full}{%
\input{expression-kinds.tex}
}{%
}%

@@

\paragraph{Managing stack size}
}{%
Continuations may lead to unbounded stack growth in
recursive code.
}%
In implementations of functional programming languages stack
growth is avoided through \textit{tail call optimization} (TCO).
However, Clojure does not support a general form of TCO, and
CPS-transformed code that creates deeply nested calls will
easily exhaust the stack. Anglican employs a workaround called
\textit{trampolining}~\cite{B95} --- instead of inserting a
continuation call directly, the transformer always wraps the
call into a \textit{thunk}, or parameterless function. The thunk
is returned and called by the trampoline (Clojure provides
function \texttt{trampoline} for this purpose) --- this way the
computation continues, but the stack is collapsed on every
continuation call. Function \texttt{continue} implements the
wrapping and is invoked on every continuation call:
\begin{lstlisting}[style=default]
=> (continue 'cont 'value 'state)
(fn [] (cont value state))
\end{lstlisting}
Correspondingly, the full, wrapped CPS form of
\begin{lstlisting}[style=default]
(fn [x y] (+ x y))
\end{lstlisting}
becomes
\begin{lstlisting}[style=default]
(fn [cont $state x y]
  (fn []
    (cont (+ x y) $state)))
\end{lstlisting}
When the CPS-transformed function is called, it returns a
\textit{thunk} (a parameterless function) which is then
re-invoked through the trampoline, with the stack collapsed.

\iftoggle{full}{%
\input{code-explosion.tex}
}{%
}%

@@

To overcome the above problems, Anglican introduces a new
addressing scheme which is almost as efficient as the scheme
based on stack addresses for reuse of previously drawn values,
while producing addresses of constant size, and allows manual
computation of checkpoint addresses at runtime when the default
automatic scheme is insufficient. According to the scheme:
\begin{itemize}
    \item A checkpoint may accept an auxiliary argument --- the
        checkpoint identifier. If specified, the identifier is the first
        argument of a checkpoint. For example \texttt{(sample
        'x1  (normal 0 1))} defines a \texttt{sample} checkpoint
        with identifier \texttt{x1}.
    \item If the identifier is omitted, a unique identifier is
        generated automatically as a fresh symbol.
    \item At runtime, the address of a checkpoint invocation has
        the form
        \texttt{[}\textit{checkpoint-identifier
        number-of-previous-occurrences}\texttt{]}, where the
        occurrences are of a checkpoint with the same checkpoint
        identifier.
    \item If a sequence of invocations of the same checkpoint is
        interrupted by a different checkpoint, the number of
        previous occurrences is rounded up to a multiple of
        an integer. For efficiency, a small power of 2 is used,
        such as $2^4 = 16$.
\end{itemize}

@@
